---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---

# Importing & Cleaning Data 

```{r warning=FALSE, warning=FALSE}
library(baseballr)
library(tidyverse)
```




```{r warning=FALSE, message=FALSE}
# all pitchers who pitched at least 10 innings for the Rays in 2020

Snell2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 605483)
Glasnow2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 607192) 
Yarbrough2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 642232)
Morton2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 450203)
Fleming2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 676596)
Fairbanks2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 664126)
Thompson2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 657044)
Slegers2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 642083)
Curtiss2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 595928)
Loup2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 571901)
Castillo2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 650895)
Beeks2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 656222)
Anderson2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 623433)
Chirinos2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 630023)
Drake2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 543118)
Richards2020 <- scrape_statcast_savant_pitcher_all(start_date="2020-07-21", end_date="2020-09-27", playerid = 670950)


TB2020 <- rbind(Snell2020, Glasnow2020, Yarbrough2020, Morton2020, Fleming2020, Fairbanks2020,Thompson2020,Slegers2020,Curtiss2020,Loup2020,Castillo2020,Beeks2020,Anderson2020,Chirinos2020,Drake2020,Richards2020)
```

```{r}
TB2020 <- TB2020 %>%
  dplyr::select(-batter, -pitcher,-spin_dir,-spin_rate_deprecated,-break_angle_deprecated,-break_length_deprecated)

TB2020 <- TB2020 %>%
  dplyr::select(-game_type,-sz_top, -sz_bot,-game_pk,-pitcher_1, -fielder_2_1, -fielder_3, -fielder_4, -fielder_5, -fielder_6, -fielder_7, -fielder_8, -fielder_9)

TB2020 <- TB2020 %>%
  dplyr::select(-tfs_deprecated, -tfs_zulu_deprecated)

TB2020 <- TB2020 %>%
  dplyr::select(-sv_id, -umpire,)

TB2020 <- TB2020 %>%
  dplyr::select(-game_year,-woba_value, -woba_denom,-babip_value,-iso_value, -pitch_number,-at_bat_number)


# https://baseballsavant.mlb.com/csv-docs#estimated_woba_using_speedangle - docs with descriptors of each variable

str(TB2020)

```

```{r warning=FALSE, message=FALSE}
for (i in 1:dim(TB2020)[1]) {
  if (TB2020$home_team[i] == "TB") {
    TB2020$opponent[i] = TB2020$away_team[i]
    TB2020$location[i] = "home"
  } else {
    TB2020$opponent[i] = TB2020$home_team[i]
    TB2020$location[i] = "away"
  }
}

for (i in 1:dim(TB2020)[1]) {
  if (is.na(TB2020$on_1b[i]) & is.na(TB2020$on_2b[i]) & is.na(TB2020$on_3b)) {
    TB2020$bases_empty[i] = 1
  } else {
    TB2020$bases_empty[i] = 0
  }
  
  if (!is.na(TB2020$on_2b[i]) || !is.na(TB2020$on_3b[i])) {
    TB2020$scoring_position[i] = 1
  } else {
    TB2020$scoring_position[i] = 0
  }
}
# create opponent variable
# create home/away variable
# make on base variable
```

```{r}
TB2020$score_margin = TB2020$fld_score - TB2020$bat_score


TB2020 <- TB2020 %>%
  dplyr::select(-pitch_type,-home_team, -away_team,-on_3b, -on_2b, -on_1b,-inning_topbot, -home_score,     -away_score, -bat_score, -fld_score, -post_away_score, -post_home_score, -post_bat_score, -post_fld_score)


for (i in 1:dim(TB2020)[1]) {
  if (TB2020$hit_location[i] == "null") {
    TB2020$hit_location[i] = NA
  }
  if (TB2020$bb_type[i] == "null") {
    TB2020$bb_type[i] = NA
  }
  if (TB2020$estimated_ba_using_speedangle[i] == "null") {
    TB2020$estimated_ba_using_speedangle[i] = NA
  }
  if (TB2020$estimated_woba_using_speedangle[i] == "null") {
    TB2020$estimated_woba_using_speedangle[i] = NA
  }
}
```

```{r}
TB2020$player_name <- as.factor(TB2020$player_name)
TB2020$stand <- as.factor(TB2020$stand)
TB2020$p_throws <- as.factor(TB2020$p_throws)
TB2020$type <- as.factor(TB2020$type)
TB2020$hit_location <- as.numeric(TB2020$hit_location)
TB2020$bb_type <- as.factor(TB2020$bb_type)
TB2020$estimated_ba_using_speedangle <- as.numeric(TB2020$estimated_ba_using_speedangle)
TB2020$estimated_woba_using_speedangle <- as.numeric(TB2020$estimated_woba_using_speedangle)
TB2020$launch_speed_angle <- as.factor(TB2020$launch_speed_angle)
TB2020$pitch_name <- as.factor(TB2020$pitch_name)
TB2020$if_fielding_alignment <- as.factor(TB2020$if_fielding_alignment)
TB2020$of_fielding_alignment <- as.factor(TB2020$of_fielding_alignment)
TB2020$opponent <- as.factor(TB2020$opponent)
TB2020$location <- as.factor(TB2020$location)
```


```{r warning=FALSE}
for (i in 1:dim(TB2020)[1]) {
  if (TB2020$events[i] == "null") {
    TB2020$events[i] = NA
  }
}

TB2020$events <- as.factor(TB2020$events)
TB2020$description <- as.factor(TB2020$description)

unique(TB2020$events)
```

```{r}
# remove pitch location outliers
TB2020 <- TB2020 %>%
  filter(plate_x < 3 & plate_x > -3 & plate_z < 6 & plate_z > -2.4)

head(TB2020)
```

# Visualizing the Data


```{r}
library(ggplot2)

ggplot(TB2020, mapping = aes(x=release_spin_rate, y=release_speed,color=pitch_name)) +
  geom_point() +
  ggtitle("Pitch Speed v. Spin Rate By Pitch Type in All 2020 Rays Pitches")
  
```

```{r}
TB2020$type <- factor(TB2020$type, levels=c("S", "B", "X"))

ggplot(TB2020, mapping = aes(x=release_spin_rate, y=release_speed,color=type)) +
  geom_point() +
  ggtitle("Pitch Speed v. Spin Rate By Pitch Outcome in All 2020 Rays Pitches")
```

```{r}

ggplot(TB2020, mapping=aes(x=plate_x, y=plate_z, color=type)) +
  geom_point() +
  ggtitle("Pitch Location by Pitch Outcome from Catcher's Perspective")
```

```{r}
ggplot(TB2020, mapping=aes(x=plate_x, y=plate_z, color=pitch_name)) +
  geom_point() +
  ggtitle("Pitch Location by Pitch Type from Catcher's Perspective")
```


# Logistic Regression on whether or not a ball is put into play

```{r}
summary(TB2020$type)

TB2020$fielder_2 <- factor(TB2020$fielder_2)
#summary(TB2020$in_play)
```

```{r}
# want to determine what factors are important in determining whether or not a ball is put into play, then can try out more advanced methods of predicting across 3+ classes

# creating factor variable indicating whether or not a ball is put into play
TB2020$in_play = NA
for (i in 1:dim(TB2020)[1]) {
  if (TB2020$type[i] == "X") {
    TB2020$in_play[i] = "1"
  } else {
    TB2020$in_play[i] = "0"
  }
}
TB2020$in_play = as.factor(TB2020$in_play)



# using cross validation - split into training and test data

set.seed(123)
train <- sample.int(7622, size=3811)

TB2020.train <- TB2020[train,]
TB2020.test <- TB2020[-train,]
```

```{r}
# select the variables that may be potentially good predictors of in_play

TB2020.inplay.train <- TB2020.train %>%
  dplyr::select(in_play, release_speed, release_pos_x, release_pos_z, player_name,zone, stand, p_throws, balls, strikes, outs_when_up, inning, fielder_2, release_spin_rate, release_extension, pitch_name, opponent, location, bases_empty, scoring_position, score_margin)


TB2020.inplay.test <- TB2020.test %>%
  dplyr::select(in_play, release_speed, release_pos_x, release_pos_z, player_name,zone, stand, p_throws, balls, strikes, outs_when_up, inning, fielder_2, release_spin_rate, release_extension, pitch_name, opponent, location, bases_empty, scoring_position, score_margin)
```

```{r}
# performing stepwise logistic regression to see which predictors should be included in the final model 
TB2020.inplay.train <- na.omit(TB2020.inplay.train)
TB2020.inplay.test <- na.omit(TB2020.inplay.test)

library(MASS)
inplaymodel <- glm(in_play ~.+release_spin_rate*pitch_name+release_speed*pitch_name, data = TB2020.inplay.train, family = binomial) %>%
  stepAIC(trace = FALSE)
# Summarize the final selected model
summary(inplaymodel)
```

```{r}
probabilities <- inplaymodel %>% predict(TB2020.inplay.test, type = "response")
predicted.classes <- ifelse(probabilities > 0.5, "1", "0")
# Model accuracy
table(predicted.classes, TB2020.inplay.test$in_play)
mean(predicted.classes==TB2020.inplay.test$in_play)
```

# LDA on whether or not a ball is put into play 

```{r}
inplaylda <- lda(in_play~release_pos_z+zone+balls+strikes+pitch_name, data=TB2020.inplay.train)

ldapredict <- predict(inplaylda, TB2020.inplay.test)

table(ldapredict$class, TB2020.inplay.test$in_play)

mean(ldapredict$class == TB2020.inplay.test$in_play)
```

```{r}
# trying with different set of predictors 

inplaylda <- lda(in_play~release_pos_z+zone+balls+strikes+release_speed + release_spin_rate + stand, data=TB2020.inplay.train)
ldapredict <- predict(inplaylda, TB2020.inplay.test)
inplaylda
table(ldapredict$class, TB2020.inplay.test$in_play)

mean(ldapredict$class == TB2020.inplay.test$in_play)

# models seem to consistently predict most pitches as being not in play
```

# Trying tree methods to predict whether or not a ball is put into play

```{r}
library(randomForest)

rf <- randomForest(
  in_play ~ .,
  data=TB2020.inplay.train
)


rfpred = predict(rf, TB2020.inplay.test)

table(rfpred, TB2020.inplay.test$in_play)

mean(rfpred == TB2020.inplay.test$in_play)
```

also tried doing random forest with different subsets of predictors, but could not find one that increased correct classification rate



# P2

# looking at only the balls that are put into play to see which factors are important in determining whether a hit occurs or not 

```{r}
# 1298 pitches were put into play against Rays Pitching in 2020
TB2020play <- TB2020 %>%
  filter(type == "X")

#str(TB2020play)

#summary(TB2020play$events)

# creating indicator variable to indicate whether or not a given ball in play resulted in a hit

TB2020play$hit = NA
for (i in 1:dim(TB2020play)[1]) {
  if (TB2020play$events[i] == "double" || TB2020play$events[i] == "single" || TB2020play$events[i] == "triple" || TB2020play$events[i] == "home_run") {
    TB2020play$hit[i] = 1
  } else {
    TB2020play$hit[i] = 0
  }
}

TB2020play$hit <- factor(TB2020play$hit)

summary(TB2020play$hit)
```

```{r}
# create test and train for in play dataset
set.seed(123)
train <- sample.int(1298, size=649)
TB2020play.train <- TB2020play[train,]
TB2020play.test <- TB2020play[-train,]
# select relevant variables for predicting hit

str(TB2020play.train)

TB2020play.test.1 <- TB2020play.test %>%
  dplyr::select(release_speed, release_pos_x, release_pos_z, player_name, zone, stand, p_throws, balls, strikes, outs_when_up, inning, release_spin_rate, release_extension, pitch_name, if_fielding_alignment, of_fielding_alignment,opponent, location, bases_empty, scoring_position, score_margin, hit)

TB2020play.train.1 <- TB2020play.train %>%
  dplyr::select(release_speed, release_pos_x, release_pos_z, player_name, zone, stand, p_throws, balls, strikes, outs_when_up, inning, release_spin_rate, release_extension, pitch_name, if_fielding_alignment, of_fielding_alignment,opponent, location, bases_empty, scoring_position, score_margin, hit)
```

# Stepwise Logistic Regression

```{r}
# performing stepwise logistic regression 
TB2020play.train.1 <- na.omit(TB2020play.train.1)
TB2020play.test.1 <- na.omit(TB2020play.test.1)

library(MASS)
hitmodel <- glm(hit~.+balls*strikes+pitch_name*release_speed+release_spin_rate*pitch_name+stand*p_throws, data = TB2020play.train.1, family = binomial) %>%
  stepAIC(trace = FALSE)
# Summarize the final selected model
summary(hitmodel)
```
```{r}
probabilities <- hitmodel %>% predict(TB2020play.test.1, type = "response")
predicted.classes <- ifelse(probabilities > 0.5, "1", "0")
# Model accuracy

table(predicted.classes, TB2020play.test.1$hit)
mean(predicted.classes==TB2020play.test.1$hit)

# model does not predict a single pitch to have a hit probability above 0.5

# will lowering the threshold to predict as a hit decrease the missclassification rate?

predicted.classes <- ifelse(probabilities > 0.45, "1", "0")
# Model accuracy

table(predicted.classes, TB2020play.test.1$hit)
mean(predicted.classes==TB2020play.test.1$hit)


predicted.classes <- ifelse(probabilities > 0.4, "1", "0")
# Model accuracy

table(predicted.classes, TB2020play.test.1$hit)
mean(predicted.classes==TB2020play.test.1$hit)



predicted.classes <- ifelse(probabilities > 0.35, "1", "0")
# Model accuracy

table(predicted.classes, TB2020play.test.1$hit)
mean(predicted.classes==TB2020play.test.1$hit)

# not evident that it does
```


# LDA
```{r}
hitlda <- lda(hit~pitch_name+release_speed+release_spin_rate+outs_when_up+release_extension+balls+strikes, data=TB2020play.train.1)
hitlda
ldapredicthit <- predict(hitlda, TB2020play.test.1)
table(ldapredicthit$class, TB2020play.test.1$hit)
mean(ldapredicthit$class == TB2020play.test.1$hit)
```


# Random Forests

```{r}
library(randomForest)

rf.hit <- randomForest(
  hit ~ .,
  data=TB2020play.train.1
)
rf.hit

rfpred.hit = predict(rf.hit, TB2020play.test.1)

table(rfpred.hit, TB2020play.test.1$hit)

mean(rfpred.hit == TB2020play.test.1$hit)
```

even after trying a variety of methods, cannot find any classification models that improve on the model other than just predicting every pitch to be a non-hit

# can be a lot of variability in what results in a hit, another goal is to minimize exit velocity, which will decrease hits over time

predicting whether or not a given ball in play results in a hit based on pitch-level and game-situation did not prove to be very successful since what is defined as a hit can range from a weak infield single to a 450 foot home run. 

plot below shows that hits tend to have slightly higher exit velocities than non-hits

but also if you induce weak contact, that means you're less likely to give up damaging hits that are more likely to produce runs 
  i.e the hits that occur with lower exit velocities are much more likely to be singles than extra base hits

```{r}
TB2020play$outcome = factor(TB2020play$events)
for (i in 1:dim(TB2020play)[1]) {
  if (TB2020play$events[i] != "double" && TB2020play$events[i] != "single" && TB2020play$events[i] != "triple" && TB2020play$events[i] != "home_run") {
    TB2020play$outcome[i] = "non-hit"
  }
}


ggplot(TB2020play) +
  geom_boxplot(mapping=aes(x=hit,y=launch_speed, col=outcome)) +
  ylim(30,120) +
  ggtitle("Boxplot of Exit Velocities Between Hits and Non-Hits") +
  scale_color_discrete(name = "Ball in Play Outcome", labels = c("Double", "Home Run", "Single", "Triple", "Non-Hit"))

```

```{r}
# creating train and test dataset with all possible predictors for hit, but with exit velocity instead of hit
TB2020play.test.2 <- TB2020play.test %>%
  dplyr::select(release_speed, release_pos_x, release_pos_z, player_name, zone, stand, p_throws, balls, strikes, outs_when_up, inning, release_spin_rate, release_extension, pitch_name, if_fielding_alignment, of_fielding_alignment,opponent, location, bases_empty, scoring_position, score_margin, launch_speed)

TB2020play.train.2 <- TB2020play.train %>%
  dplyr::select(release_speed, release_pos_x, release_pos_z, player_name, zone, stand, p_throws, balls, strikes, outs_when_up, inning, release_spin_rate, release_extension, pitch_name, if_fielding_alignment, of_fielding_alignment,opponent, location, bases_empty, scoring_position, score_margin, launch_speed)
```

# Stepwise Regression on Exit Velocity

```{r}
TB2020play.train.2 <- na.omit(TB2020play.train.2)
TB2020play.test.2 <- na.omit(TB2020play.test.2)

# removing any outlier exit velocities
# many very slow exit velocities occurred in model that were likely from bunts and check swings that should not be included to fit the model

TB2020play.train.2 <- TB2020play.train.2 %>%
  filter(launch_speed > 45)

TB2020play.test.2 <- TB2020play.test.2 %>%
  filter(launch_speed > 45)

library(MASS)
exitvelomodel <- glm(launch_speed~.+balls*strikes+pitch_name*release_speed+release_spin_rate*pitch_name+stand*p_throws-if_fielding_alignment-of_fielding_alignment, data = TB2020play.train.2, family = gaussian) %>%
  stepAIC(trace = FALSE)
# Summarize the final selected model
summary(exitvelomodel)
```

```{r}
exitvelopredict <- predict(exitvelomodel, TB2020play.test.2)

mean((exitvelopredict - TB2020play.test.2$launch_speed)^2)
```

# Regularization cv.glmnet methods to predict exit velocity
```{r}
library(modelr)
TB2020play.2 <- rbind(TB2020play.test.2, TB2020play.train.2)

y = TB2020play.2$launch_speed
X=model_matrix(TB2020play.2,launch_speed~.+release_speed*release_spin_rate+stand*p_throws+zone*release_speed)[,-22]
var.names=names(X)
dim(X)
```



```{r}
library(glmnet)
set.seed(216)
cvmod.0=cv.glmnet(y=y,x=as.matrix(X),type.measure = "mse",alpha=0)
set.seed(216)
cvmod.25=cv.glmnet(y=y,x=as.matrix(X), type.measure = "mse",alpha=0.25)
set.seed(216)
cvmod.5=cv.glmnet(y=y,x=as.matrix(X), type.measure = "mse",alpha=0.5)
set.seed(216)
cvmod.75=cv.glmnet(y=y,x=as.matrix(X), type.measure = "mse",alpha=0.75)
set.seed(216)
cvmod.1=cv.glmnet(y=y,x=as.matrix(X), type.measure = "mse",alpha=1)

CV.0.ERROR=cvmod.0$cvm[which(cvmod.0$lambda==cvmod.0$lambda.1se)]
CV.25.ERROR=cvmod.25$cvm[which(cvmod.25$lambda==cvmod.25$lambda.1se)]
CV.5.ERROR=cvmod.5$cvm[which(cvmod.5$lambda==cvmod.5$lambda.1se)]
CV.75.ERROR=cvmod.75$cvm[which(cvmod.75$lambda==cvmod.75$lambda.1se)]
CV.1.ERROR=cvmod.1$cvm[which(cvmod.1$lambda==cvmod.1$lambda.1se)]

MOD.RESULT=tibble(alpha=c(0,0.25,0.5,0.75,1),
                  lambda=c(cvmod.0$lambda.min,cvmod.25$lambda.min,
                           cvmod.5$lambda.min,cvmod.75$lambda.min,
                           cvmod.1$lambda.min),
                  CV.Error=c(CV.0.ERROR,CV.25.ERROR,CV.5.ERROR,
                             CV.75.ERROR,CV.1.ERROR))
print(MOD.RESULT)
```

```{r}
best.alpha=MOD.RESULT$alpha[which.min(MOD.RESULT$CV.Error)]
best.lambda=MOD.RESULT$lambda[which.min(MOD.RESULT$CV.Error)]

best.mod=glmnet(y=y,x=as.matrix(X),nlambda=1,lambda=best.lambda,alpha=best.alpha)
best.coef=as.tibble(as.matrix(coef(best.mod)))
best.coef2=best.coef %>% 
              mutate(Parameter=c("Int",var.names)) %>%
              rename(Estimate=s0) %>%
              dplyr::select(Parameter,Estimate)
nonzero.best.coef=best.coef2 %>%
                    filter(Estimate!=0)
print(nonzero.best.coef,n=1e3)

TB2020play.2$exitvelo.hat=predict(best.mod,newx=as.matrix(X))

ggplot(TB2020play.2) +
  geom_point(aes(x=launch_speed,y=exitvelo.hat),color="lightskyblue2") +
  geom_abline(a=0,b=1,linetype="dashed") +
  theme_minimal() +
  ylab("Predicted Exit Velocity") +
  xlab("Actual Exit Velocity")

ggplot(TB2020play.2) +
  geom_histogram(aes(x=launch_speed-exitvelo.hat),fill="lightskyblue2") +
  theme_minimal() +
  xlab("Residuals") +
  ylab("Frequency")


# residuals not very normally distributed & best model only includes zone, which can't be good for prediction stepwise regression model is better than this
```


```{r}
library(kableExtra)
kable(nonzero.best.coef) %>% kable_styling('condensed')
```

```{r}
#write.csv(TB2020, file="TBRaysPitchers2020.csv")
```

